---
bibliography: papers.bib
---

```{r}
#| include: false

library(here)
library(DiagrammeR)

```

# Estimation of effects using causal mediation analysis
<!-- Daniel 120 min -->

## Recap from yesterday

- causal inference in general
- causal mediation analysis
  - the controlled direct effect (CDE)
  - natural direct effect (NDE)
  - natural indirect effect (NIE)

```{r, engine="tikz", fig.cap = "Fig. 3. DAG under *no intermediate confounders* of the mediator-outcome relation affected by treatment", fig.ext=if (knitr:::is_latex_output()) "pdf" else "png", echo = FALSE}
\dimendef\prevdepth=0
\pgfdeclarelayer{background}
\pgfsetlayers{background,main}
\usetikzlibrary{arrows,positioning}
\tikzset{
>=stealth',
punkt/.style={
rectangle,
rounded corners,
draw=black, very thick,
text width=6.5em,
minimum height=2em,
text centered},
pil/.style={
->,
thick,
shorten <=2pt,
shorten >=2pt,}
}
\newcommand{\Vertex}[2]
{\node[minimum width=0.6cm,inner sep=0.05cm] (#2) at (#1) {$#2$};
}
\newcommand{\VertexR}[2]
{\node[rectangle, draw, minimum width=0.6cm,inner sep=0.05cm] (#2) at (#1) {$#2$};
}
\newcommand{\ArrowR}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) to[bend right=30] (#2);
\end{pgfonlayer}
}
\newcommand{\ArrowL}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) to[bend left=45] (#2);
\end{pgfonlayer}
}
\newcommand{\EdgeL}[3]
{ \begin{pgfonlayer}{background}
\draw[dashed,#3] (#1) to[bend right=-45] (#2);
\end{pgfonlayer}
}
\newcommand{\Arrow}[3]
{ \begin{pgfonlayer}{background}
\draw[->,#3] (#1) -- +(#2);
\end{pgfonlayer}
}
\begin{tikzpicture}
  \Vertex{-4, 0}{W}
  \Vertex{0, 0}{M}
  \Vertex{-2, 0}{A}
  \Vertex{2, 0}{Y}
  \Arrow{W}{A}{black}
  \Arrow{A}{M}{black}
  \Arrow{M}{Y}{black}
  \ArrowL{W}{Y}{black}
  \ArrowL{A}{Y}{black}
  \ArrowL{W}{M}{black}
\end{tikzpicture}
```

## Regression-based approach

### Controlled direct effect

### Natural direct effect

### Natural indirect effect

## Proportion eliminated

### 2-way decomposition of effects

### R package example for causal mediation analysis - Mediation-package

## G-computation-based approach

### Controlled direct effect

### Natural direct effect

### Natural indirect effect

## 3-way and 4-way decomposition of effects

### R package example for causal mediation analysis - CMAverse-package


