# Survival outcomes

<!-- Teaching concepts - Jie (30 min) -->

::: callout-note
## Learning outcomes

-   Perform causal mediation analysis with survival outcomes using R
    software
:::

There are many studies conducting mediation analyses with
**time-to-event outcomes**. Survival analysis allows investigators to
study these important outcomes with appropriate consideration for
variable follow-up times, censoring, and competing risks.

Take the red meat example for instance, we previously demonstrated
higher red meat intake increased blood glucose. However, we might be
even more interested in whether red meat intake increase the onset of
type 2 diabetes (T2D). If we have the diagnosis date or duration of T2D,
we can conduct survival analyses on red meat and T2D onset.

Another example is the effect of risk factor on patient survival time. A
study investigated the effect of socioeconomic status (SES) on the
survival time of cancer patients and how much cancer stage mediated the
effect. They found the effect of SES on survival time was partially
mediated by stage diagnosis, explaining 12% for lung cancer.

```{r echo=FALSE}
# Creating The causal diagram for a mediation model
library(DiagrammeR)
grViz("
digraph {
  graph []
  node [shape = plaintext]
    X [label = 'socioeconomic status']
    M [label = 'stage diagnosis']
    Y [label = 'cancer mortality']
  edge [minlen = 1.5]
    X->Y
    X->M
    M->Y
  { rank = same; X; M; Y }
}")
```

### Conduct causal mediation analysis for time-to-event outcomes

How to conduct causal mediation analysis for time-to-event outcomes? We
introduced the difference and product methods for continuous and binary
outcomes in previous session. It is tempting to run a linear regression
model for the mediator and proportional hazard model for the outcome,
then use product or difference method to estimate the direct effect and
indirect effect.

#### Product method for time-to-event outcomes

The Cox proportional hazards model is commonly used for dealing with
survival data in medical literature. Cox regression estimates the
**hazard ratios** and the values are then used to determine the effect
of the mediator variable between the exposure and the survival time of
outcome.

Could we use the traditional approach for time-to-event outcomes? We
have introduced the difference and product methods for continuous and
binary outcomes in previous session. It is tempting to run a linear
regression model for the mediator and proportional hazard model for the
outcome, then use product or difference method to estimate the direct
effect and indirect effect.

However, 'non-collaspsibility' is a problem of the hazard ratio as odds
ratio (VanderWeele). Therefore, use of Cox PH regression to
approximately estimate indirect effects via difference or product of
coefficients rests on the assumption that the outcome is **rare**
(VanderWeele).

Where the outcome is common, measures of the indirect effect or
proportion mediated will be incorrect. Tein and Mackinnon (2003)
considered whether the product method and difference method yield
comparable results with respect to time-to-event outcomes. They found
that the methods coincides for the accelerated failure time model but
not for the proportional hazards model.

To sum up, we can only use the traditional approaches for rare outcomes.
Otherwise, we can use the product method to get an indication of whether
there is mediation, but be aware that the estimate is not accurate.

### Causal mediation for time-to-event outcomes

::: callout-note
In earlier session,we have been familiar with the counterfactual
concepts in causal mediation:

-   $Yi(a, m)$ is the outcome achieved for person i if, possibly
    contrary to fact, exposure had been set to a and mediator to m.

-   $Mi(a)$ is the mediator achieved for person i if, possibly contrary
    to fact, exposure had been set to a.

One can combine the two counterfactuals, yielding so-called nested
counterfactuals defined as

-   $Y(a, M (a\*)$.

By introducing the nested counterfactual for a ≠ a\* we can give a
precise mathematical definition of mediation.
:::

For a **survival outcome**, the outcome of interest will be survival
time (SV).

-   $SV (t) = P(V ≥ t)$ the survival function at time t

-   $SV (t\|c)=P(V ≥ t\|c)$ the survival function conditional on
    covariates C

-   $λV (t)$ : the hazard at time t

-   $λV (t\|c)$: conditional hazard at time t

### Examples

::: callout-note We will continue working on the obesity-CVD example in
the Framingham dataset.

The outcome of interest is death from cardiovascular diseases (cvd). The
underlying time scale is in days, starting at participants entered the
cohort. The exposure of interest is obesity status at baseline, where
a=1 indicates obese, a=0 indicates non-obese. The mediator is blood
pressure, the counterfactual mediator was M(a), where M(a) is the
mediator when the exposure is obesity (1) and M(0) is the mediator
values when the exposure is not-obese.

The question of interest here is whether the blood pressure mediates the
impact of obesity on CVD-related death (measured in years).

```{r}
library(tidyverse)
framingham <- read.csv(here::here("data/framingham_dataset.csv"))

framingham <- framingham %>%
  select(
    id = randid,
    bmi,
    m = sysbp,
    y = cvd,
    w1 = sex,
    w2 = cursmoke,
    w3 = age,
    y_time = timecvd
  ) %>%
  na.omit() %>%
  mutate(
    a = case_when(
      bmi >= 25 ~ 1,
      TRUE ~ 0
    ),
    y_time = y_time / 365.25
  ) 
```

The new thing here is that we need to calculate the time to event. This
is already done in the dataset.

We need the outcome to be the time to event and specify event as the
outcome variable. Then we use the package as before.

```{r}
library(CMAverse)
library(survival)

res_rb_coxph <- cmest(
  data = framingham, model = "rb", outcome = "y_time", event = "y", exposure = "a",
  mediator = "m", basec = c("w1", "w2", "w3"), EMint = TRUE,
  mreg = list("linear"), yreg = "coxph",
  astar = 0, a = 1, mval = list(120),
  estimation = "imputation", inference = "bootstrap"
)


summary(res_rb_coxph)
```
