# Survival outcomes

<!-- Teaching concepts - Jie (30 min) -->

<!-- Daniel will look and trim down a bit - Jie (30 min) -->

<!-- Coding part - Daniel Teaching concepts - Jie (30 min)  -->

## Time-to-event outcomes

There are many studies conducting mediation analyses with
**time-to-event outcomes**. Survival analysis allows researchers to
study these outcomes while considering follow-up times, censoring, and
competing risks.

Take the red meat study for instance, we previously demonstrated higher
red meat intake increased blood glucose. However, we might be even more
interested in whether red meat intake increase the onset of type 2
diabetes (T2D). If we have the diagnosis date or duration of T2D, we can
conduct survival analyses on red meat and T2D onset.

Another example is the effect of risk factor on patient survival time. A
study investigated the effect of socioeconomic status (SES) on the
survival time of cancer patients and how much cancer stage mediated the
effect. They found the effect of SES on survival time was partially
mediated by stage diagnosis, explaining 12% for lung cancer.

```{r echo=FALSE}
# Creating The causal diagram for a mediation model
library(DiagrammeR)
grViz("
digraph {
  graph []
  node [shape = plaintext]
    X [label = 'socioeconomic status']
    M [label = 'stage diagnosis']
    Y [label = 'cancer mortality']
  edge [minlen = 1.5]
    X->Y
    X->M
    M->Y
  { rank = same; X; Y }
  { rank = min; M;}
}")
```

### Conduct causal mediation analysis for time-to-event outcomes

#### Product method for time-to-event outcomes

The Cox proportional hazards model is commonly used for dealing with
survival data in medical literature. Cox regression estimates the
**hazard ratios** and the values are then used to determine the effect
of the mediator variable between the exposure and the survival time of
outcome.

Could we use the traditional approach for time-to-event outcomes? We
have introduced the difference and product methods for continuous and
binary outcomes in previous session. It is tempting to run a linear
regression model for the mediator and proportional hazard model for the
outcome, then use product or difference method to estimate the direct
effect and indirect effect.

However, 'non-collaspsibility' is a problem of the hazard ratio as odds
ratio (VanderWeele). Therefore, use of Cox PH regression to
approximately estimate indirect effects via difference or product of
coefficients rests on the assumption that the outcome is **rare**
(VanderWeele).

Where the outcome is common, measures of the indirect effect or
proportion mediated will be incorrect. Tein and Mackinnon (2003)
considered whether the product method and difference method yield
comparable results with respect to time-to-event outcomes. They found
that the methods coincides for the accelerated failure time model but
not for the proportional hazards model.

To sum up, we can only use the traditional approaches for rare outcomes.
Otherwise, we can use the product method to get an indication of whether
there is mediation, but be aware that the estimate is not accurate.

#### Causal mediation for time-to-event outcomes

So how to conduct causal mediation analysis for time-to-event outcomes?

::: callout-note
In earlier session,we have been familiar with the counterfactual
concepts in causal mediation:

-   $Yi(a, m)$ is the outcome achieved for person i if, possibly
    contrary to fact, exposure had been set to a and mediator to m.

-   $Mi(a)$ is the mediator achieved for person i if, possibly contrary
    to fact, exposure had been set to a.

One can combine the two counterfactuals, yielding so-called nested
counterfactuals defined as

-   $Y(a, M (a\*)$.

By introducing the nested counterfactual for a ≠ a\* we can give a
precise mathematical definition of mediation.
:::

Once you have grasped the core builds on comparing distributions of
nested counterfactuals, these effects can just as easily be expressed on
other scales than the averages.

For a **survival outcome**, the outcome of interest will be survival
time (SV).

-   $SV (t) = P(V ≥ t)$ the survival function at time t

-   $SV (t\|c)=P(V ≥ t\|c)$ the survival function conditional on
    covariates C

-   $λV (t)$ : the hazard at time t

-   $λV (t\|c)$: conditional hazard at time t

#### Definitions

If we consider the survival functions for a time-to-event outcome T, we
could decompose the survival function as follows:

$ST_a(t) - ST_a*(t) = [ST_aM_a(t)-ST_aM_a*(t)] + [ST_aM_a*(t)-ST_a*M_a*(t)]$

The first expression in brackets is the natural indirect effect on the
survival function scale and the second is the natural direct effect on
the survival function scale.

Similarly, we can demcompose the overal difference in hazards on the
hazard scale:
$λT_a(t) - λT_a*(t) = [λT_aM_a(t)-λT_aM_a*(t)] + [λT_aM_a*(t)-λT_a*M_a*(t)]$

### Assumptions of mediation analysis with a time-to-event outcome

Similar as our context, mediation analysis with a time-to-event outcome
have to satisfy below assumptions:

-   no unmeasured confounding of the exposure-outcome relation;

-   no unmeasured confounding of the mediator-outcome relation;

-   no unmeasured confounding of the exposure-mediation relation;

-   no exposure induced mediation-outcome confounding

-   Additionally, we assume that the mediator is measured for everyone
    before the outcome occurs.

When the outcome is common, we can use weight approach (Lange 2012)

### Examples

::: callout-note
We will continue working on the obesity-CVD example in the Framingham
dataset. The outcome of interest is death from cardiovascular diseases
(cvd). The underlying time scale is in days, starting at participants
entered the cohort. - The exposure of interest is obesity status at
baseline, where a=1 indicates obese, a=0 indicates non-obese.

-The mediator is blood pressure, the counterfactual mediator was M(a),
where M(a) is the mediator when the exposure is obesity (1) and M(0) is
the mediator values when the exposure is not-obese.

-The nested counterfactual was death time. For ease of exposition, we
will only refer to the underlying event time and event indicator when
required by context.
:::

The question of interest here is whether the blood pressure mediates the
impact of obesity on CVD-related death (measured in years).

::: callout-note
## Learning outcomes

-   Perform causal mediation analysis using R software
-   Interpret results from causal mediation analysis
:::

-   survival analysis
-   code-along with package and setup of survival data

<!-- Coding part - Daniel -->

```{r}
library(tidyverse)
framingham <- read.csv(here::here("data/framingham_dataset.csv"))

framingham <- framingham %>%
  select(
    id = randid,
    bmi,
    m = sysbp,
    y = cvd,
    w1 = sex,
    w2 = cursmoke,
    w3 = age,
    y_time = timecvd
  ) %>%
  na.omit() %>%
  mutate(
    a = case_when(
      bmi >= 25 ~ 1,
      TRUE ~ 0
    ),
    y_time = y_time / 365.25
  ) 
```

> Does obesity lead to higher CVD risk by increasing blood pressure?

The new thing here is that we need to calculate the time to event. This
is already done in the dataset.

We need the outcome to be the time to event and specify event as the
outcome variable.

```{r}
library(CMAverse)
library(survival)

res_rb_coxph <- cmest(
  data = framingham, model = "rb", outcome = "y_time", event = "y", exposure = "a",
  mediator = "m", basec = c("w1", "w2", "w3"), EMint = TRUE,
  mreg = list("linear"), yreg = "coxph",
  astar = 0, a = 1, mval = list(120),
  estimation = "imputation", inference = "bootstrap"
)


summary(res_rb_coxph)
```
