# Day 1: Group work

```{r}
library(here)
```

For the exercises, we will use the Framingham dataset.

```{r}
framingham <- read.csv(here::here("data", "frmgham2.csv"))
```

The overall aim of the cohort study was to investigate effect of obesity
on the risk of cardiovascular disease (CVD).

## Variables

The dataset includes 'r nrow(framingham)´ participants and 'r
ncol(framingham)´.

```{r}
str(framingham)
```

For the exercise on day 1, we will work with the following variables:

-   socio = socioeconomic status (SES) (binary: 0 = low SES, 1 = high
    SES)
-   smoke = smoking status (binary: 0 = no prenatal smoking, 1 =
    prenatal smoking)
-   educ = education ()
-   bmi = body mass index (BMI) (continuous variable in kg/m^2^)
-   CVD= (binary: 0=no, 1=yes).
-   obesity (binary: 0=normal, 1=obesity)\*

There are other variables in the dataset, please take a few minutes to
explore the dataset. Before we start, let's think about the research
question and draw a directed acyclic graphs. You can either draw by hand
or use daggity (https://www.dagitty.net/).

[Address the following questions]{.underline}:

1.  Estimate the conditional treatment effect of obesity on the
    prevalence of CVD (conditional on SES, or other variables based on
    DAG).

2.  Estimate the relationship between obesity and blood pressure.

3.  Fit the outcome model, with obesity as exposure, and blood pressure
    as outcome.

4.  Calculate the direct effect and indirect effect using the product
    method and difference method, and compare results from the two
    approaches.

5.  Calculate the direct effect and indirect effect using the
    **mediation** package.

6.  Discuss the limitations of the traditional approach.
