# Day 2: Group work

::: callout-note
# Objective of this session

In this session, We will get hands on practice using the framingham
dataset.

-   Conduct sensitivity analyses

-   Conduct mediation analyses with multiple mediators
:::

For the exercises, we will continue using the Framingham dataset. From
the group work yesterday, we have learned how to conduct a mediation
analyses. But in real life, the research question could be more complex.
After running the main mediation analyses, how robust the mediation
effect is? What is there are more than 1 mediator you are interested in?

```{r}
library(here)
```

```{r}
framingham <- read.csv(here::here("data", "frmgham2.csv"))
```

We will keep working the obesity and cardiovascular disease example.
First, think about the group work yesterday, and reflect the following
questions:

-   go over the assumptions of the causal mediation, and discuss how
    robust your estimates are

-   are there interactions between obesity and hypertension?

-   are there other mediators other than systolic blood pressure?

## Variables

The dataset includes 'r nrow(framingham)´ participants and 'r
ncol(framingham)´.

```{r}
str(framingham)
```

For the exercise on day 1, we will work with the following variables:

-   socio = socioeconomic status (SES) (binary: 0 = low SES, 1 = high
    SES)
-   smoke = smoking status (binary: 0 = no prenatal smoking, 1 =
    prenatal smoking)
-   educ = education ()
-   bmi = body mass index (BMI) (continuous variable in kg/m^2^)
-   CVD= (binary: 0=no, 1=yes).
-   obesity (binary: 0=normal, 1=obesity)\*

There are other variables in the dataset, please take a few minutes to
explore the dataset. Before we start, let's think about the research
question and draw a directed acyclic graphs. You can either draw by hand
or use daggity (https://www.dagitty.net/).

[Address the following questions]{.underline}: You can either use the
**mediation** package or **CMAverse** package.

1.  Re-do the DAG, and investigate all the potential pathways

2.  Conduct a sensitivity analyses and discuss the results

3.  Write a report of the study
