# Day 2: Group work

::: callout-note
# Objective of this session

In this session, We will get hands on practice using the framingham
dataset.

-   Conduct sensitivity analyses

-   Conduct multiple mediators

-   
:::

For the exercises, we will continue using the Framingham dataset. From
the group work yesterday, we have learned how to conduct a mediation
analyses. But in real life, the research question could be more complex.
After running the main mediation analyses, how robust the mediation
effect is? What is there are more than 1 mediator you are interested in?

```{r}
library(here)
```

```{r}
framingham <- read.csv(here::here("data", "frmgham2.csv"))
```

We will keep working the obesity and cardiovascular disease example.
First, think about the group work yesterday, and reflect the following
questions:

-   go over the assumptions of the causal mediation, and discuss how
    robust your estimates are

-   are there interactions between obesity and hypertension?

-   are there other mediators other than systolic blood pressure?

## Variables

The dataset includes 'r nrow(framingham)´ participants and 'r
ncol(framingham)´.

```{r}
str(framingham)
```

For the exercise on day 1, we will work with the following variables:

-   socio = socioeconomic status (SES) (binary: 0 = low SES, 1 = high
    SES)
-   smoke = smoking status (binary: 0 = no prenatal smoking, 1 =
    prenatal smoking)
-   educ = education ()
-   bmi = body mass index (BMI) (continuous variable in kg/m^2^)
-   CVD= (binary: 0=no, 1=yes).
-   obesity (binary: 0=normal, 1=obesity)\*

There are other variables in the dataset, please take a few minutes to
explore the dataset. Before we start, let's think about the research
question and draw a directed acyclic graphs. You can either draw by hand
or use daggity (https://www.dagitty.net/).

[Address the following questions]{.underline}:

1.  Estimate the conditional treatment effect of obesity on the
    prevalence of CVD (conditional on SES, or other variables based on
    DAG).

2.  Estimate the relationship between obesity and blood pressure.

3.  Fit the outcome model, with obesity as exposure, and blood pressure
    as outcome.

4.  Calculate the direct effect and indirect effect using the product
    method and difference method, and compare results from the two
    approaches.

5.  Calculate the direct effect and indirect effect using the
    **mediation** package.

6.  Discuss the limitations of the traditional approach.
