# Day 2: Group work

<!-- conduct causal mediation analysis, interpretation, conduct sensitivity analysis, interpretation, reporting according to guidelines - Jie will take first look  -->

::: callout-note
# Objective of this session

In this session, We will get hands on practice using the framingham
dataset.

-   Conduct sensitivity analyses

-   Conduct mediation analyses with multiple mediators
:::

```{r}
library(here)
```

```{r}
framingham <- read.csv(here::here("data", "frmgham2.csv"))
```

We will keep working the obesity and cardiovascular disease example.
First, think about the group work yesterday, and reflect the following
questions with your group member.

-   go over the assumptions of the causal mediation, and discuss how
    robust your estimates are

-   are there interactions between obesity and hypertension?

-   are there other mediators other than systolic blood pressure?

[Address the following questions]{.underline}: You can either use the
**mediation** package or **CMAverse** package.

1.  Re-do the DAG, and investigate all the potential pathways

2.  Conduct a sensitivity analyses and discuss the results

3.  Write a report of your mediation models
