# Day 2: Group work

<!-- conduct causal mediation analysis, interpretation, conduct sensitivity analysis, interpretation, reporting according to guidelines - Jie will take first look  -->

::: callout-note
# Objective of this session

In this session, We will get hands on practice using the framingham
dataset. You will conduct a proper causal mediation analysis,
considering:

-   Conduct mediation analyses with exposure induced mediator-outcome
    confounder or multiple mediators

-   Conduct sensitivity analyses

-   Report the results following the guidelines
:::

```{r}
library(here)
```

```{r}
framingham <- read.csv(here::here("data", "frmgham2.csv"))
```

We will continue working the obesity and cardiovascular disease example.
First, reflect on the group work from yesterday and discuss the
following questions with your group members:

-   Review the assumptions of causal mediation, and discuss the
    robustness of your estimates

-   Are there other mediators other than systolic blood pressure?

-   Are there interactions between obesity and the mediators?

Re-conduct the mediation analysis and address the discussion points. You
can either use the **mediation** package or **CMAverse** package.

1.  Re-do the DAG, and investigate all the potential pathways

2.  Conduct a sensitivity analyses and discuss the results

3.  Write a report of your mediation models
