# Motivation for mediation analysis

<!-- Omar 30 min -->

```{r}
#| include: false

# Creating The causal diagram for a mediation model
library(DiagrammeR)
```

Basic, clinical and epidemiological research are often interested in
whether a specific intervention has an effect on a disease. Once this is
established, the natural subsequent question is **why** this
intervention had an effect? [@vanderweele_mediation_2016].

The subsequent and most relevant question is **HOW?**

Mediation analysis answers the question regarding **HOW** a
treatment/exposure/intervention impacts an outcome of interest by
uncovering the underlying pathways, and mechanisms.

## Motivation for mediation analysis

### Explanation and understanding

```{r}
#| echo: false
library(DiagrammeR)
grViz("
digraph {
  graph []
  node [shape = plaintext]
    X [label = 'Family history of diabetes']
    M [label = 'genetic makeup']
    N [label = 'shared environment']
    Y [label = 'Type 2 diabetes']
  edge [minlen = 2]
    X->M
    M->Y
    X->N
    N->Y
{ rank = same; M; X; Y; }
  { rank = min; N;}
}
")
```

### Confirmation and refutation of theory

<!-- DBI: this section is a little unclear. There are 2 mediators in the example above. -->

Let's assume we estimate a familial risk of diabetes and that risk
estimates follow a normal distribution at the population level. Then we
select the extremes of the distribution (2.5%) and assess the relative
contribution of the genetic makeup, and risk phenotypes, among
individuals with a strong/weak familial risk of type 2 diabetes.

### To refine interventions

Let's consider the evidence from the landmark prevention trials. The
Diabetes Prevention Programme (DPP, USA), The Diabetes Prevention Study
(DPS, Finland), and The Da Quin study (DPS, China). These studies
randomized high risk individuals to either metformin or a lifestyle
intervention consisting of a physical activity or physical activity +
diet. The main finding was a \~58% lower risk of incidence type 2
diabetes in the lifestyle intervention group compared to the metformin
group

```{r}
#| echo: false

# Creating The causal diagram for a mediation model
library(DiagrammeR)
grViz("
digraph {
  graph []
  node [shape = plaintext]
    X [label = 'Lifestyle intervention']
    M [label = 'Weight loss']
    Y [label = 'Type 2 diabetes']
  edge [minlen = 2]
    X->M
    M->Y
   { rank = same; M; X; Y; }
}
")
```

We might be interested in further refining the intervention so as to
increase the magnitude of the effect. This might be done by altering or
improving components of the intervention that target a particular
mechanism for the outcome and discarding those components of which
perhaps are not relevant for the outcome

When implementing the same lifestyle interventions in real world
settings, lifestyle interventions show smaller effect and the long-term
adoption and sustainability over time were some of the challenges to the
effectiveness of these prevention efforts

```{r}
#| echo: false

# Creating The causal diagram for a mediation model
library(DiagrammeR)
grViz("
digraph {
  graph []
  node [shape = plaintext]
    X [label = 'Lifestyle intervention']
    M [label = 'Weight loss']
    N [label = '?']    
  Y [label = 'Type 2 diabetes']
  edge [minlen = 2]
    X->M
    M->Y
    N->M
   {rank = same; M; X; Y; }
   {rank = min; N; }
}
}
")
```

## Motivations for running this course

### Methods for conducting mediation analysis are used incorrectly

A recent systematic review of the reporting quality of studies applying
mediation analysis showed that most of the studies had incomplete
reporting [@rrn_systematic_2022]. This might impact both reproducibility
and translation into practice.

### The methodological framework for conducting mediation analysis evolved rapidly in recent years

Therefore, most applied examples are outdated in mediation analysis
methods. A scoping review found that between 2015 to 2019 only 29.3% of
studies using mediation analysis used modern mediation analysis methods
[@rijnhart_mediation_2021].

::: callout-note
## Before moving on to the first exercise

Spend 5 minutes talking to your neighbour and discuss your own
motivations for participating in this course. Consider the different
motivations behind conducting mediation analysis.
:::

::: callout-note
## Exercise: "Understanding different study designs". Duration: 30 minutes

You will be divided into small groups of four people. Your assignment
consist on reading the selected mediation analyses sections of the
assigned articles and briefly present in plenum the methodological
approach of your paper, and share the more relevant points discussed in
your group in 3 minutes. Each paper has a different study design and you
will have 15 min to complete the reading task described below.

### Group 1: Intervention study a

Paper: [Shoer et al.
2023](https://www.nature.com/articles/s41467-023-41042-x)

-   read the abstract
-   read section 'Mediation analysis' under 'Methods'
-   read section 'The microbiome mediates the diet's effect' under
    'Results'
-   try to understand Fig. 4.

Task: explain the overall study and the mediation analysis. What does
the results imply?

### Group 2: Intervention study b

Paper: [Johansen et al.
2020](https://link.springer.com/article/10.1007/s00125-020-05249-7)

-   read the abstract
-   read from "Single mediation analysis" under 'Statistical analysis'
-   try to understand table 4
-   read from "To explore the role of weight loss in relation to beta
    cell function we performed single mediation analysis" under
    'Discussion'

Task: explain the overall study and the mediation analysis. What does
the results imply? How did the authors interpret the results? How do you
interpret the results?

### Group 3: Cohort study

Paper: [Wang et al. 2022](https://doi.org/10.1007/s00125-022-05692-8)

-   read the abstract
-   read from "To test whether an association between the metabolite
    profile score and risk of type 2 diabetes..." under 'Statistical
    analysis'
-   try to understand Figure 4
-   read the conclusion

Task: explain the overall study and the mediation analysis. What does
the results imply? How did the authors interpret the results? How do you
interpret the results?

### Group 4: Cross-sectional study

Paper: [Shi et al.
2022](https://www.nature.com/articles/s41598-022-08544-y)

-   read the abstract
-   they used a specific algorithm to identify mediation. You do not
    have to understand how it works. Simply note whether they adjusted
    for specific variables in their analysis (Hint: look in the methods
    section)
-   Read section 'Mediation by the gut microbial species Dialister
    invisus' under 'Results'

Task: explain the overall study design, what exposure, mediator and
outcomes were they interested in? Are their results causal?
:::

## References
