# Motivation for mediation analysis

<!-- Omar 30 min -->

Basic, clinical and epidemiological research are often interested in
whether a specific intervention has an effect on a disease. Once this is
established, the natural subsequent question is **WHY** this
intervention had an effect? [@vanderweele_mediation_2016]. The
subsequent and most relevant question is **HOW?**

Mediation analysis answers the question regarding **HOW** a
treatment/exposure/intervention impacts an outcome of interest by
uncovering the underlying pathways, and mechanisms.

1- The first motivation may be simply explanation and understanding.

```{r}
# Creating The causal diagram for a mediation model
library(DiagrammeR)
grViz("
digraph {
  graph []
  node [shape = plaintext]
    X [label = 'Family history of diabetes']
    M [label = 'genetic makeup']
    N [label = 'shared environment']
    Y [label = 'Type 2 diabetes']
  edge [minlen = 2]
    X->M
    M->Y
    X->N
    N->Y
{ rank = same; M; X; Y; }
  { rank = min; N;}

}
")
```

2- Confirmation and refutation of theory Let's assume we estimate a
familial risk of diabetes and that risk estimates follow a normal
distribution at the population level. Then we select the extremes of the
distribution (2.5%) and assess the relative contribution of the genetic
makeup, and risk phenotypes, among individuals with a strong/weak
familial risk of type 2 diabetes.

3- To refine interventions Let's consider the evidence from the landmark
prevention trials. The Diabetes Prevention Programme (DPP, USA), The
Diabetes Prevention Study (DPS, Finland), and The Da Quin study(DPS,
China). These studies randomized high risk individuals to either
metformin or a lifestyle intervention consisting of a diet + physical
activity. The main finding was a \~57% relative risk reduction in the
incidence of type 2 diabetes in the lifestyle intervention group vs.
metformin group

```{r}
# Creating The causal diagram for a mediation model
library(DiagrammeR)
grViz("
digraph {
  graph []
  node [shape = plaintext]
    X [label = 'Lifestyle intervention']
    M [label = 'Weight loss']
    Y [label = 'Type 2 diabetes']
  edge [minlen = 2]
    X->M
    M->Y
   { rank = same; M; X; Y; }
}
")
```

We might be interested in further refining the intervention so as to
increase the magnitude of the effect. This might be done by altering or
improving components of the intervention that target a particular
mechanism for the outcome and discarding those components of which
perhaps are not relevant for the outcome

When implementing the same lifestyle interventions in real world
settings, lifestyle interventions show smaller effect and the long-term
adoption and sustainability over time were some of the challenges to the
effectiveness of these prevention efforts

Alongside the applied research, the methodological framework for
conducting mediation analysis has also evolved rapidly in recent years.
Unfortunately, most applied examples have not caught up with this
progression in methodology. In fact, a scoping review found that between
2015 to 2019 only 29.3% of studies using mediation analysis used modern
mediation analysis methods [@rijnhart_mediation_2021].

```{r}
# Creating The causal diagram for a mediation model
library(DiagrammeR)
grViz("
digraph {
  graph []
  node [shape = plaintext]
    X [label = 'Lifestyle intervention']
    M [label = 'Weight loss']
    N [label = '?']    
  Y [label = 'Type 2 diabetes']
  edge [minlen = 2]
    X->M
    M->Y
    X->N
    N->Y
   {rank = same; M; X; Y; }
   {rank = min; N; }
}
}
")
```

## Motivations for running this course:

-   Methods for conducting mediation analysis are used incorrectly. A
    recent systematic review of the reporting quality of studies
    applying mediation analysis showed that most of the studies had
    incomplete reporting [@rrn_systematic_2022]. This might impact both
    reproducibility and translation into practice.
-   The methodological framework for conducting mediation analysis
    evolved rapidly in recent years. Therefore, most applied examples
    are outdated in mediation analysis methods. A scoping review found
    that between 2015 to 2019 only 29.3% of studies using mediation
    analysis used modern mediation analysis methods
    [@rijnhart_mediation_2021].

::: callout-note \## Before moving on to the first exercise. Spend 5
minutes talking to your neighbour and discuss your own motivations for
participating in this course. Consider the different motivations behind
conducting mediation analysis.

::: callout-note
## Exercise: Small groups work. "Understanding different study designs". Duration:30 minutes

You will be divided into small groups of four people. Your assignment
consist on reading the selected mediation analyses sections of the
assigned articles and briefly present in plenum the methodological
approach of your paper, and share the more relevant points discussed in
your group in 3 minutes. Each of paper have a different study design and
you will have 15 min to complete the reading task described below. :::
callout-note \## Exercise: Understanding different study designs (30
min)

You will now be placed in different groups and read selected sections of
different mediation analyses within different study designs. You have 15
min to complete the task. Then, each group will briefly present their
paper, the approach used and what they thought about it.

### Group 1: Intervention study a

Paper: [Shoer et al.
2023](https://www.nature.com/articles/s41467-023-41042-x)

-   read the abstract
-   read section 'Mediation analysis' under 'Methods'
-   read section 'The microbiome mediates the diet’s effect' under
    'Results'
-   try to understand Fig. 4.

Task: explain the overall study and the mediation analysis. What does
the results imply?

### Group 2: Intervention study b

Paper: [Johansen et al.
2020](https://link.springer.com/article/10.1007/s00125-020-05249-7)

-   read the abstract
-   read from "Single mediation analysis" under 'Statistical analysis'
-   try to understand table 4
-   read from "To explore the role of weight loss in relation to beta
    cell function we performed single mediation analysis" under
    'Discussion'

Task: explain the overall study and the mediation analysis. What does
the results imply? How did the authors interpret the results? How do you
interpret the results?

### Group 3: Cohort study

Paper: [Wang et al. 2022](https://doi.org/10.1007/s00125-022-05692-8)

-   read the abstract
-   read from "To test whether an association between the metabolite
    profile score and risk of type 2 diabetes..." under 'Statistical
    analysis'
-   try to understand Fig. 4
-   read the conclusion

Task: explain the overall study and the mediation analysis. What does
the results imply? How did the authors interpret the results? How do you
interpret the results?

### Group 4: Cross-sectional study

Paper: [Shi et al.
2022](https://www.nature.com/articles/s41598-022-08544-y)

-   read the abstract
-   they used a specific algorithm to identify mediation. You do not
    have to understand how it works. Simply note whether they adjusted
    for specific variables in their analysis (Hint: look in the methods
    section)
-   Read section 'Mediation by the gut microbial species Dialister
    invisus' under 'Results'

Task: explain the overall study design, what exposure, mediator and
outcomes were they interested in? Are their results causal?
:::

## References
