# Traditional mediation analysis

```{r setup}
#| include: false
library(here)
library(DiagrammeR)
library(tidyverse)
library(dplyr)
```

::: callout-note
## Objective of this session

-   To learn how to conduct mediation analysis using the traditional
    approaches.
-   Understanding the assumptions and limitations of the methods.
:::

The purpose of mediation analysis is to determine if the effect of a
treatment (A) on an outcome (Y) can be explained by a third mediating
variable (M). Thus, mediation analysis not only answers whether two
variables are related, but also **why**.

This can be visualized in the following figures.

Overall relationship between A and Y:

```{r}
#| echo: false

grViz("
digraph {
  graph []
  node [shape = plaintext]
    A [label = 'A']
    Y [label = 'Y']
  edge [minlen = 1.5]
    A->Y [label = 'c']
  { rank = same; A; Y }
}")
```

Relationship between A and Y through M:

```{r echo=FALSE}
#| echo: false

grViz("
digraph {
  graph []
  node [shape = plaintext]
    A [label = 'A']
    M [label = 'M']
    Y [label = 'Y']
  edge [minlen = 1.5]
    A->Y [label = 'c*', xlabel = 'c']
    A->M [taillabel = 'a']
    M->Y [label = 'b']
  { rank = same; A; Y }
  { rank = min; M;}
}")
```

## Traditional approaches for mediation analysis

The two traditional approaches to mediation analysis are **the
difference method** and **the product method**; also known as the Baron
& Kenny-method @baron_moderatormediator_1986.

The traditional approach use these equations for mediator and outcome
(for the case of a continuous mediator and a continuous outcome)

Mediator-model:

$E(M|A=a, C=c) = \beta_0 + \beta_1a + \beta_2c$ (3.1)

Outcome model with adjustment for mediator:

$E(Y|A=a, M=m, C=c) = \theta_0 + \theta_1a + \theta_2m + \theta4c$ (3.2)

Outcome model without adjustment for mediator:

$E(Y|A=a, C=c) = \theta_0' + \theta_1'a + \theta4'c$ (3.3)

Note:

-   total effect=$\theta_1'$ , that is the total effect of the
    independent variable on the dependent variable

-   direct effect=$\theta_1$', that is the effect of the independent
    variable on the dependent variable that is not mediated by the
    mediator

-   mediation effect=$\beta1*\theta_2$ (the product method)

-   mediation effect='$\theta_1'-\theta_1$ (the difference method)

Note: The terms 'mediation effect' and 'indirect effect' are used
synonymously.

## An example

In this session, we will use an example to address a mediation research
question using both traditional approaches.

It has been suggested that red meat consumption is associated with
higher blood glucose levels and an increased risk of Type 2 diabetes
(T2D). However, the mechanisms are poorly understood. One hypothesis is
that red meat intake affects the level of inflammation, which in turn
influence blood glucose.

Does red meat intake directly influence blood glucose levels, or does it
act through intermediary mechanisms, such as inflammation? Let's imagine
there exists a hypothetical inflammatory biomarker that serves as a
mediator between red meat consumption and blood glucose levels.
Therefore, our interest is now how much the effect of red meat on blood
glucose is mediated by this inflammatory biomarker. The situation is
depicted by below figure.

> Does greater consumption of red meat lead to elevated levels of blood
> glucose by mediating specific inflammation biomarkers?

```{r echo=FALSE}
#| echo: false

grViz("
digraph {
  graph []
  node [shape = plaintext]
    X [label = 'Red meat']
    M [label = 'Inflammation']
    Y [label = 'Blood glucose']
  edge [minlen = 1.5]
    X->Y [label = 'c*', xlabel = 'c']
    X->M [taillabel = 'a']
    M->Y [label = 'b']
  { rank = same; X; Y }
  { rank = min; M;}
}")
```

::: callout-note
## Example Dataset: A Brief Overview

We'll explore the relationship between red meat intake and glucose
levels in the National Health and Nutrition Examination Survey (NHANES)
dataset, which has been modified for this course with simulated
variables.

To try this out in practice, we load the NHANES dataset, only keep the
variables we need for the example, remove missing data and scale the
intake of red meat to be per 65 g/day (i.e., 1-unit higher = 65 g/day).

The variables used in the dataset:

-   exposure of interest (a): red meat intake (continuous variable,
    grams per day)

-   outcome (y): plasma fasting glucose (continuous variable)

-   mediator (m): a new inflammatory biomarker (continuous variable)

Note that the example is created to demonstrate the methods, and there
is no clinical relevance of the results. For the convenience of
analysis, we have transformed the mediator (m) and outcome variable (y)
so that they follow a normal distribution.

Four demographic variables are considered in the analysis: age, gender,
education, and smoking. We assume that adjusting for these four
confounding is sufficient to block the backdoor paths. Further
discussion on confounding adjustment will be covered in the following
session.
:::

```{r}

nhanes <- read_csv(here::here("data/nhanes_dataset.csv"))

nhanes <- nhanes %>%
  select(
    id = seqn, w1 = age, w2 = gender, w3 = education_clean, w4 = smoke,
    a = total_redmeat, # this is the exposure
    m = magic_biomarker, # this is the mediator
    y = blood_glucose # this is the outcome
  ) %>% 
  na.omit()
```

## Method 1: Baron & Kenny (the product method)

According to Baron and Kenny (1986) @baron_moderatormediator_1986, the
following criteria need to be satisfied for a variable to be considered
a mediator:

1.  The exposure should be associated with the mediator.
2.  In the model for the outcome that includes both the exposure and
    mediator: the mediator should be associated with the outcome.
3.  In the model for the outcome that includes only the exposure: the
    exposure should be associated with the outcome.
4.  When controlling for the mediator, the association between the
    exposure and outcome should be reduced, with the strongest
    demonstration of mediation occurring when the path from the exposure
    to the outcome, when controlling for the mediator is zero.

The following shows the basic steps for mediation analysis suggested by
Baron & Kenny.

### Step 1: Estimate the relationship between A on M (path a)

That is to check the relation between exposure (red meat) and the
mediator (inflammatory biomarker), which corresponds to criteria 1. A
mediation makes sense only if A affects M. If A and M have no
relationship, M is just a third variable that may or may not be
associated with Y.

As the mediator is a continuous variable, we can build a linear
regression model, adjusting for confounders on the pathway between A and
M.

$(E(M|A=a, C=c) = \beta_0 + \beta_1a + \beta_2c$ (3.1)

```{r}
fita <- lm(m ~ a + w1 + w2 + w3 + w4, data = nhanes)
summary(fita)
```

Here, we see that higher red meat intake is positively associated with
this new inflammatory biomarker (an increase of 1 unit in red meat
consumption is associated with a 25.9 unit increase in the biomarker
level).

### Step 2: Estimate the relationship between M on Y controlling for A (path b)

That is to build the model between inflammatory biomarker and glucose
levels, controlling for red meat intake. This step corresponds to
criteria (ii).

$(E(Y|A=a, M=m, C=c) = \theta_0 + \theta_1a + \theta_2m + \theta4c$
(3.2)

```{r}
fitb <- lm(y ~ a + m + w1 + w2 + w3 + w4, data = nhanes)

summary(fitb)
```

After adjusting for red meat intake, the inflammatory biomarker is
associated with higher blood glucose (an increase of 1 unit in the
biomarker is associated with 0.69 unit increase in the blood glucose).

### Step 3: Estimate the relationship between A on Y (path c)

That is to build the model between red meat intake and glucose levels,
without adjusting for the inflammatory biomarker. The only difference
between step 2 and step 3 is that mediator is not included in the model.

$(E(Y|A=a, C=c) = \theta_0' + \theta_1'a + \theta4'c$ (1.3)

```{r}
fitc <- lm(y ~ a + w1 + w2 + w3 + w4, data = nhanes)
summary(fitc)
```

We can see that red meat intake is strongly associated with higher
glucose levels (an increase of 1 unit in red meat consumption is
associated with a 70.5 unit increase in the blood glucose.

Then we can estimate the direct effect and indirect effect:

-   **direct effect** = $\theta_1$\
-   **indirect effect** = $\beta_1 \theta_2$\
-   **total effect** = $\theta_1'$

The *direct effect* ($\theta_1$) of red meat on glucose levels can be
assessed by the coefficient of red meat in model b.

```{r}
direct_ma <- fitb$coefficients[2]
direct_ma
```

The *indirect effect* can be assessed by calculating the product of
$\beta_1$ $\theta_2$. In this example, the indirect effect between red
meat and glucose levels through inflammatory biomarker is 17.8.

```{r}
indirect_ma <- fita$coefficients[2] * fitb$coefficients[3]
indirect_ma
```

The *total effect* ($\theta_1'$) of red meat on glucose levels can be
assessed by the coefficient of red meat in model c, which is 70.5. You
might have noticed the total effect also equals the sum of the direct
effect and indirect effect.

```{r}
total_ma <- fitc$coefficients[2]
total_ma
```

## Method 2: Difference approach

The difference approach is more commonly used in epidemiology. And the
rational of this approach is to compare the effect of A on Y without and
with adding M in the models.

If a mediation effect exists, the effect of A on Y will be attenuated
when M is included in the regression, indicating the effect of A on Y
goes through M. If the effect of A on Y completely disappears, M fully
mediates between A and Y (full mediation). If the effect of A on Y still
exists, but in a smaller magnitude, M partially mediates between A and Y
(partial mediation).

Back to the example, the indirect effect will be the coefficient of red
meat in model c (without adjusting for m) minus the coefficient of red
meat in model b (adjusting for m), which equals to 17.8.

```{r}
indirect_mb <- fitc$coefficients[2] - fitb$coefficients[2]
indirect_mb
```

::: callout-note
The algebraic equivalence of the indirect effect using the product
method and the difference method will coincide for a continuous outcome
on the difference scale (MacKinnon 1995). However, the two methods
diverge when using a binary outcome and logistic regression. (MacKinnon
and Dwyer 1993).
:::

::: callout-note
Recall the four criteria suggested by Baron and Kenny. One of the
criteria is that 'Path C must be significantly different from 0 to
ensure there is a total effect between X and Y'. However, this step can
be controversial. Even if we don't find a significant association
between X and Y, we could proceed to the next step if we have a strong
theoretical basis for their relationship.
:::

::: callout-note
## Group work - 15 minutes

-   Discuss with your group about the two traditional approaches, did
    you get the same results?
-   Discuss the scenarios under which there might be no significant
    association between A and Y.
:::

### Use mediation packages

Now we have estimated the direct and indirect effect using the product
and the difference methods.You might have noticed what you got so far
are only point estimates, but what if you want to see if this mediation
effect is statistically significant (different from zero or not)?

To do so, there are two main approaches: the Sobel test
@sobel_asymptotic_1982 and bootstrapping @preacher_spss_2004.

Here we will demonstrate that we can use the mediate() function in
'mediation' package @tingley_mediation_2014 to conduct mediation
analysis. One of the very good options to conduct mediation analysis
using the mediation package is we can get confidence interval by
bootstrapping.

Let's load up the R packages.

```{r}
library(mediation) # Mediation package
```

mediate() takes two model objects as input ($A \rightarrow M$ and
$A + M \rightarrow Y$) and we need to specify which variable is an IV
(treatment) and a mediator (mediator). For bootstrapping, set boot =
TRUE and sims to at least 500.

```{r}
set.seed(260524)

# Fit mediator model
lm_m <- lm(m ~ a + w1 + w2 + w3 + w4, data = nhanes)
# Fit outcome model
lm_y <- lm(y ~ a + m + w1 + w2 + w3 + w4, data = nhanes)
# Run mediation analysis
results <- mediate(lm_m, lm_y, treat = "a", mediator = "m", boot = TRUE, sims = 10) #can set to 500

summary(results)
```

-   **ADE** stands for average direct effects. It describes the direct
    effect of the A on the Y when controlling for the mediator. Note
    this estimate is the same as the coefficient We got in the product
    and difference method. But be aware we are able to get *confidence
    interval* and *significance* levels for the indirect effect, not
    only its two individual parts by bootstrapping. This is something we
    need for reporting the mediation.

-   **ACME** stands for average causal mediation effects. This is the
    indirect effect of the IV on the DV that goes through the mediator.
    Note that ACME estimated to be 17.8, which is the same to the
    traditional approached we just used.

-   **Total Effect** stands for the total effect (direct + indirect) of
    the IV onto the DV. We can also get it by adding the ACME () and the
    ADE () to receive the total effect.

-   **Prop. Mediated** describes the proportion of the effect of the IV
    on the DV that goes through the mediator. It's calculated by
    dividing the ACME (17.8) by the total effect 70.5) and yields 0.252.
    The result suggests the mediator (inflammatory biomarker) explains
    25.3% proportion of the effect of red meat on glucose levels. We
    will talk more about proportion mediation in the next session.

```{r}
# We can also plot the estimates
plot(results)
```

For details of the mediate() package, please refer to
@tingley_mediation_2014.

## Limitations of the traditional approach

-   Non-linearity
-   Interactions
-   Multiple mediators

```{r}
#| echo: false

grViz("
digraph {
  graph []
  node [shape = plaintext]
    X [label = 'Red meat']
    M [label = 'Inflammation']
    Y [label = 'Blood glucose']
    N[label = 'Insulin sensitivity']
    A[label = 'Fasting insulin']
  edge [minlen = 1.5]
    X->Y 
    X->M 
    M->Y 
    X->N    
    N->Y
    X->A    
    A->Y
  { rank = same; X; N; Y; }
  { rank = min; M;}
  { rank = max; A;}
}")
```

## When can we use the traditional approach

When fulfill the criteria, simple tools like regression can be used to
estimate a causal mediation effect:

-   no unmeasured confounding
-   no exposure-mediator interaction
-   linear relationship
-   rare binary outcome

## References
